Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-23 paper/main.tex
backing up changes:
- finished addressing R1 comments
- addressed some of R2's comments

% supplemental tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}

% supplemental figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\individualKnowledgeMapsA}{3}
\newcommand{\individualKnowledgeMapsB}{4}
\newcommand{\individualKnowledgeMapsC}{5}
\newcommand{\individualLearningMapsA}{6}
\newcommand{\individualLearningMapsB}{7}

@@

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics. Conceptually, increasing the
number of topics used by the model functions to increase the ``resolution'' of
the embedding space, providing a greater ability to estimate knowledge for
content that is highly similar to (but not precisely the same as) that
contained in the two lectures. Aside from increasing the number of topics from
15 to 100, all other procedures and model parameters were carried over from the
preceding analyses. As in our other analyses, we resampled each lecture's topic
trajectory to 1~Hz and projected each question into a shared text embedding
space.

@@

We view our work as making several contributions to the study of how people
acquire conceptual knowledge. First, from a methodological standpoint, our
modeling framework provides a systematic means of mapping out and
characterizing knowledge in maps that have infinite (arbitrarily many) numbers
of coordinates, and of ``filling out'' those maps using relatively small
numbers of multiple choice quiz questions. Our experimental finding that we can
use these maps to predict responses to held-out questions
(Fig.~\ref{fig:predictions}) also has important psychological implications. One
such psychological implication is that concepts that are assigned to nearby
coordinates by the text embedding model also appear to be ``known to a similar
extent'' (as reflected by participants' responses to held-out questions). This
suggests that participants also \textit{conceptualize} similarly the content
reflected by nearby embedding coordinates. A second psychological implication
is that something about how participants' knowledge ``spreads'' across concepts
is being captured by the knowledge maps we are inferring from their quiz
responses (e.g., Fig.~\ref{fig:smoothness}). In other words, our study shows
that knowledge about a given concept implies knowledge about related concepts.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces}
(i.e., Lecture~1), 15 about the conceptual content of \textit{Birth of Stars}
(i.e., Lecture~2), and 9 questions that tested for general conceptual knowledge
about basic physics (covering material that was not presented in either video).
We note that estimating the specific ``amount'' of conceptual understanding
that each question ``requires'' to answer is somewhat subjective, and might
even come down to the ``strategy'' a given participant uses to answer the
question at that particular moment. The full set of questions and answer
choices may be found in Supplementary Table~\questions.

@@

We fetched these automated transcripts using the
\texttt{youtube-transcript-api} Python package~\citep{Depo18}. The transcripts
consisted of one timestamped line of text for every few seconds (mean: 2.34~s;
standard deviation: 0.83~s) of spoken content in the video (i.e., corresponding
to each individual caption that would appear on-screen if viewing the lecture
via YouTube, and when those lines would appear). We defined a sliding window
length of (up to) $w = 30$ transcript lines, and assigned each window a
timestamp corresponding to the midpoint between the timestamps for its first
and last lines. These sliding windows ramped up and down in length at the
beginning and end of each transcript, respectively. In other words, each
transcript's first sliding window covered only its first line, the second
sliding window covered the first two lines, and so on. This ensured that each
line from the transcripts appeared in the same number ($w$) of sliding windows.
After performing various standard text preprocessing (e.g., normalizing case,
lemmatizing, removing punctuation and stop-words), we treated the text from
each sliding window as a single ``document,'' and combined these documents
across the two videos' windows to create a single training corpus for the topic
model.

@@

Intuitively, $\mathrm{ncorr}(x, y)$ is the correlation between two topic
vectors (e.g., the topic vector from one timepoint in a lecture, $x$, and the
topic vector for one question, $y$), normalized by the minimum and maximum
correlations (across all timepoints $t$ and questions $Q$) to range between 0
and 1, inclusive. Equation~\ref{eqn:prop} then computes the weighted average
proportion of correctly answered questions about the content presented at
timepoint $t$, where the weights are given by the normalized correlations
between timepoint $t$'s topic vector and the topic vectors for each question.
The normalization step (i.e., using $\mathrm{ncorr}$ instead of the raw
correlations) insures that every question contributes some non-negative amount
to the knowledge estimate.

>>>>

% supplemental tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}
\newcommand{\matchTab}{3}

% supplemental figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\individualKnowledgeMapsA}{3}
\newcommand{\individualKnowledgeMapsB}{4}
\newcommand{\individualKnowledgeMapsC}{5}
\newcommand{\individualLearningMapsA}{6}
\newcommand{\individualLearningMapsB}{7}
\newcommand{\ldaVsBERT}{8}
\newcommand{\jaccard}{9}   % word match figure

@@

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics. Conceptually, increasing the
number of topics used by the model functions to increase the ``resolution'' of
the embedding space, providing a greater ability to estimate knowledge for
content that is highly similar to (but not precisely the same as) that
contained in the two lectures. This change in the number of topics overcame an
undesireable behavior in the UMAP embedding procedure~\citep{McInEtal18},
whereby embedding coordinates for the 15-topic model tended to be ``clumped''
into separated clusters, rather than forming a smooth trajectory through the 2D
space. When we increased the number of topics to 100, the embedding coordinates
in the 2D space formed a smooth trajectory through the space, with
substantially less clumping (Fig.~\ref{fig:knowledge-maps}). We note that we
used these 2D maps solely for visualization; all relevant comparisons, distance
computations, and statistical tests we report above were carried out in the
original 15-dimensional space, using the 15-topic model. Aside from increasing
the number of topics from 15 to 100, all other procedures and model parameters
were carried over from the preceding analyses. As in our other analyses, we
resampled each lecture's topic trajectory to 1~Hz and projected each question
into a shared text embedding space.

@@

We view our work as making several contributions to the study of how people
acquire conceptual knowledge. First, from a methodological standpoint, our
modeling framework provides a systematic means of mapping out and
characterizing knowledge in maps that have infinite (arbitrarily many) numbers
of coordinates, and of ``filling out'' those maps using relatively small
numbers of multiple choice quiz questions. Our experimental finding that we can
use these maps to predict responses to held-out questions
(Fig.~\ref{fig:predictions}) also has important psychological implications. One
such psychological implication is that concepts that are assigned to nearby
coordinates by the text embedding model also appear to be ``known to a similar
extent'' (as reflected by participants' responses to held-out questions). This
suggests that participants also \textit{conceptualize} similarly the content
reflected by nearby embedding coordinates. A second psychological implication
is that something about how participants' knowledge ``spreads'' across concepts
is being captured by the knowledge maps we are inferring from their quiz
responses (e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In
other words, our study shows that knowledge about a given concept implies
knowledge about related concepts.

In our study, we characterize the ``coordinates'' of participants' knowledge
using a relatively simple ``bag of words'' text embedding model~\citep[LDA;
]{BleiEtal03}. More sophisticated text embedding models, such as
transformer-based models~\citep{ViswEtal17, DevlEtal18, ChatGPT, TouvEtal23}
can learn complex grammatical and semantic relationships between words,
higher-order syntactic structures, sylistic features, and more. We considered
using transformer-based models in our study, but we were surprised to find that
the text embeddings derived from these models were surprisingly uninformative
with respect to differentiating or otherwise characterizing the conceptual
content of the lectures and questions we used. We suspect that this reflects a
broader challenge in constructing models that are high-resolution within a
given domain (e.g., the domain of physics lectures and questions) \textit{and}
sufficiently broad so as to enable them to cover a wide range of domains. For
example, we found that the embeddings derived even from much larger and more
modern models like BERT~\citep{DevlEtal18}, GPT~\citep{ViswEtal17},
LLaMa~\citep{TouvEtal23}, and others that are trained on enormous text corpora,
end up yielding poor resolution within the content space spanned by indiviudal
course videos (Supp. Fig.~\ldaVsBERT). Whereas the LDA embeddings of the
lectures and questions are ``near'' each other (i.e., the convex hull enclosing
the two lectures' trajectories is highly overlapping with the convex hull
enclosing the questions' embeddings), the BERT embeddings of the lectures and
questions are instead largely distinct (top row of Supp. Fig.~\ldaVsBERT). The
LDA embeddings of the questions for each lecture and the corresponding
lecture's trajectory are also similar. For example, as shown in
Fig.~\ref{fig:sliding-windows}C, the LDA embeddings for \textit{Four
Fundamental Forces} questions (blue dots) appear closer to the \textit{Four
Fundamental Forces} lecture trajectory (blue line), whereas the LDA embeddings
for \textit{Birth of Stars} questions (green dots) appear closer to the
\textit{Birth of Stars} lecture trajectory (green line). The BERT embeddings of
the lectures and questions do not show this property (Supp. Fig.~\ldaVsBERT).
We also examined per-question ``content matches'' between individual questions
and individual moments of each lecture
(Figs.~\ref{fig:question-correlations},~\ldaVsBERT). The timeseries plot of
individual questions' correlations are different from each other when computed
using LDA (e.g., the traces can be clearly visually separated), whereas the
correlations computed from BERT embeddings of different questions all look very
similar. This tells us that LDA is capturing some differences in content
between the questions, whereas BERT is not. The timeseries plots of individual
questions' correlations have clear ``peaks'' when computed using LDA, but not
when computed using BERT. This tells us that LDA is capturing a ``match''
between the content of each question and a relatively well-defined time window
of the corresponding lectures. The BERT embeddings appear to blur together the
content of the questions versus specific moments of each lecture. Finally, we
also compared the pairwise correlations between embeddings of questions within
versus across content areas (i.e., content covered by the individual lectures,
lecture-specific questions, and by the ``general physics knowledge''
questions). The LDA embeddings show a strong contrast between same-content
embeddings versus across-content embeddings. In other words, the embeddings of
questions about the \textit{Four Fundamental Forces} material are highly
correlated with the embeddings of the \textit{Four Fundamental Forces} lecture,
but not with the embeddings of \textit{Birth of Stars}, questions about
\textit{Birth of Stars}, or general physics knowledge questions. We see a
similar pattern with the LDA embeddings of the \textit{Birth of Stars}
questions (Fig.~\ref{fig:topics}, Supp. Fig.~\topicWeights). In contrast, the
BERT embeddings are all highly correlated with each other (Supp.
Fig.~\ldaVsBERT). Taken together, these comparisons illustrate how LDA (trained
on the specific content in question) provides both coverage of the requisite
material and specificity at the level of the content covered by individual
questions. BERT, on the other hand, essentially assigns both lectures and all
of the questions (which are all broadly about ``physics'') into a tiny region
of its embedding space, thereby blurring out meaningful distinctions between
different specific concepts covered by the lectures and questions. We note that
these are not criticisms of BERT (or other large language models trained on
large and diverse corpora). Rather, our point is that simple fine-tuned models
trained on a relatively small but specialized corpus can outperform much more
complicated models trained on much larger corpora, when we are specifically
interested in capturing subtle conceptual differences at the level of a single
course lecture or question. Of course if our goal had been to find a model that
generalized to many different content areas, we would expect our approach to
perform comparatively poorly relative to BERT or other much larger models. We
suggest that bridging the tradeoff between high resolution within each content
area versus the ability to generalize to many different content areas will be
an important challenge for future work in this domain.

At the opposite end of the spectrum from large language models, one could also
imagine \textit{simplifying} some aspects of our LDA-based approach by
computing simple word overlap metrics. For example, the Jaccard similarity
between text $A$ and $B$ is computed as the number of unique words in the
intersection of words from $A$ and $B$ divided by the number of unique words in
the union of words from $A$ and $B$. In a supplemental analysis (Supp.
Fig.~\jaccard), we compared the LDA-based question-lecture matches we reported
in Figure~\ref{fig:question-correlations} with the Jaccard similarities between
each question and each sliding window of text from the corresponding lecture.
As shown in Supplementary Figure~\jaccard, this simple word-matching approach
does not appear to capture the same level of specificity as the LDA-based
approach. For example, whereas the LDA-based approach often yields a clear peak
in the timeseries of correlations between each question and the corresponding
lecture, the Jaccard similarity-based approach does not. Furthermore, these
LDA-based matches appear to capture conceptual overlaps between the questions
and lectures (Supp. Tab.~\matchTab), whereas simple word matching does not. For
example, one of the example questions examined in Supplementary Figure~\jaccard
asks ``Which of the following occurs as a cloud of atoms gets more dense?''.
The LDA-based matches identify lecture timepoints where the releveant
\textit{topics} are discussed (e.g., when words like ``cloud,'' ``atom,''
``dense,'' etc., are mentioned \textit{together}). The Jaccard similarity-based
matches, on the other hand, are strong when \textit{any} of these words are
mentioned, even if they do not occur together.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). All participants viewed the videos in the
same order (i.e., \textit{Four Fundamental Forces} followed by \textit{Birth of
Stars}). While we are not aware of any specific confounds of viewing order, nor
have are we aware of how or why viewing order might influence our main
findings, we acknowledge that we did not control for potential order effects in
our study.

We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces}
(i.e., Lecture~1), 15 about the conceptual content of \textit{Birth of Stars}
(i.e., Lecture~2), and 9 questions that tested for general conceptual knowledge
about basic physics (covering material that was not presented in either video).
One of our group's undergraduate research assistants worked alongside a
rotating Masters student to develop this set of questions (these researchers
are acknowledged in our paper for their contribution, although they did not
meet the criteria for authorship discussed with all team members at the start
of the project, as determined by J.R.M.) The senior author (J.R.M.) tasked the
pair of researchers with coming up with ``15 conceptual questions about each
lecture, along with 9 additional questions about general physics knowledge.'' To
help broaden the set of lecture-specific questions, the researchers were
further instructed to work through each lecture in small segments, identify
what each segment was ``about'' conceptually, and then write a question about
that concept. The general physics questions were drawn from the researchers'
coursework along with internet searches and brainstorming with the project team
and other members of J.R.M.'s lab. The final set of questions (and response
options) was reviewed and approved by J.R.M.

We note that estimating the specific ``amount'' of conceptual understanding
that each question ``requires'' to answer is somewhat subjective, and might
even come down to the ``strategy'' a given participant uses to answer the
question at that particular moment. The full set of questions and answer
choices may be found in Supplementary Table~\questions.

@@

We fetched these automated transcripts using the
\texttt{youtube-transcript-api} Python package~\citep{Depo18}. The transcripts
consisted of one timestamped line of text for every few seconds (mean: 2.34~s;
standard deviation: 0.83~s) of spoken content in the video (i.e., corresponding
to each individual caption that would appear on-screen if viewing the lecture
via YouTube, and when those lines would appear). After scraping the
transcripts, we carried out several standard preprocessing steps. First, we
removed all punctuation, replaced numbers with their word equivalents (e.g.,
``1'' with ``one''), and converted all text to lowercase. Next, we removed all
stop words, as defined by the Natural Language Toolkit~\citep[NLTK;
]{BirdEtal09} English stop word list, augmented with the following additional
words that we hand-identified as frequently occuring and non-diagnostic:
``actual,'' ``actually,'' ``also,'' ``bit,'' ``could,'' ``e,'' ``even,''
``first,'' ``follow,'' ``following,'' ``four,'' ``let,'' ``like,'' ``mc,''
``really,'', ``saw,'' ``see,'' ``seen,'' ``thing,'' and ``two.'' We defined a
sliding window length of (up to) $w = 30$ transcript lines, and assigned each
window a timestamp corresponding to the midpoint between the timestamps for its
first and last lines. This $w$ parameter was chosen to match the same number of
words per sliding window (rounded to the nearest whole word, and after
preprocessing) as the sliding windows we defined in our prior
work~\citep{HeusEtal21}. The sliding windows contained an average of 73.8 words
and lasted for an average of 62.22 seconds.

These sliding windows ramped up and down in length at the beginning and end of
each transcript, respectively. In other words, each transcript's first sliding
window covered only its first line, the second sliding window covered the first
two lines, and so on. This ensured that each line from the transcripts appeared
in the same number ($w$) of sliding windows. After performing various standard
text preprocessing (e.g., normalizing case, lemmatizing, removing punctuation
and stop-words), we treated the text from each sliding window as a single
``document,'' and combined these documents across the two videos' windows to
create a single training corpus for the topic model.

@@

Intuitively, $\mathrm{ncorr}(x, y)$ is the correlation between two topic
vectors (e.g., the topic vector from one timepoint in a lecture, $x$, and the
topic vector for one question, $y$), normalized by the minimum and maximum
correlations (across all timepoints $t$ and questions $Q$) to range between 0
and 1, inclusive. Equation~\ref{eqn:prop} then computes the weighted average
proportion of correctly answered questions about the content presented at
timepoint $t$, where the weights are given by the normalized correlations
between timepoint $t$'s topic vector and the topic vectors for each question.
The normalization step (i.e., using $\mathrm{ncorr}$ instead of the raw
correlations) ensures that every question contributes some non-negative amount
to the knowledge estimate.
